To be FAIR: Theory Specification Needs an Update

### 2025-06-07 action_letter.Rmd
update 2025-06-07 16:43:02.696576

>>>>

To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
We also already have two worked examples for actual theories, which we now point out in the text (with links to the relevant vignettes).

### 2025-06-07 test.Rmd
update 2025-06-07 16:43:02.696576

To concretize the FAIR principles,
we propose an applied workflow for making theory FAIR.
The guiding principle of our approach is to align and build upon existing successful open science infrastructures to the greatest possible extent.
At the time of writing (2024),
the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* FAIR principles.
Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
a popular platform in psychology.
Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory amongst users of that platform,
while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
Note that open science infrastructure is an area of active development, and as such,
the approach proposed here might change as new tools or databases are developed or existing tools and database change over time.
The following conceptual workflow does not require the use of R, but most steps can be automated in R using the `theorytools` package.
The package also includes a living document with the [latest version of the workflow](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
We present a brief summary of the workflow at the time of writing here,
to illustrate the general principles of FAIRifying theory which can also be implemented using other open science infrastructures.

@@

If we compare this to the levels of theory formalization [@guestHowComputationalModeling2021],
it is defined at either the "theory" or "specification" level.
We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
The implementation below describes the model as a directed graph (see also Figure \@ref(fig:figecs)a).
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
We can save this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

```
digraph {

  observation;
  induction;
  deduction;
  test;
  evaluation;

}
```

## 2. Creating a Project Folder

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html).
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

<!-- ### What's in a README? -->

<!-- The readme should contain information to help people get started with using your FAIR theory. -->
<!-- We suggest the following elements: -->

<!-- * Title, prefaced with `# FAIR theory: The Theory's Name` -->
<!-- * Description: A plain-text description of the theory and its scope -->
<!-- * Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use". From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)". We propose explicitly addressing the theory's X-Interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph. -->
<!-- * Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*. -->
<!-- * License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*. -->
<!-- * Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page. -->
<!-- * Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here: -->

<!-- ``` -->
<!-- This repository contains an implementation of the "empirical cycle", -->
<!-- a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works. -->

<!-- > De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology: -->
<!-- Foundations of inference and research in the behavioral sciences. -->
<!-- De Gruyter Mouton. https://doi.org/10.1515/9783112313121 -->
<!-- ``` -->

## 3. Version Control the Repository

<!-- The field of computer science provides well-established processes for creating information artifacts that can be iteratively improved. -->
<!-- In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021]. -->
To version control our project, we initiate a Git repository in the project folder.
<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
<!-- opening a terminal in the project folder, -->
<!-- and running the command `git init`. -->
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`. -->
<!-- The default settings for the repository are fine; -->
<!-- for the next steps of the tutorial it  -->
Note that the repository
must be set to "Public" to take advantage of GitHub's Zenodo integration.
Push the local files to the Git remote repository, and keep them synchronized going forward.

@@

First, create a Zenodo account with your existing GitHub account.
Then in Zenodo, go to the GitHub section under your account.
Following the instructions on the page, activate Zenodo for your theory repository.
Then, create a new release of the GitHub repository.
Choose a tag and release title using our adapted semantic versioning, starting with version 1.0.0, if you intend to share your theory with the broader scientific community.
After publishing the release,
you should be able to see the archived version in your Zenodo account,
along with a DOI.

@@

To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle, as archived in Zenodo.

- Set the *resource type* to `Model`; this ensures proper archival in Zenodo
- Verify that the *title* is prefaced with `FAIR theory:`; this allows sentient readers to recognize the work as a FAIR theory
- Add the *keyword* `fairtheory`; this aids search engine indexation
- Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
- List the DOIs/identifiers of *related works*. Use the `Relation` field as appropriate. For example:
    + `Is documented by` can be used to reference a theory paper you wrote, in which you introduce this FAIR theory
    + `Is derived from` could be used to reference a paper or book chapter that introduced an existing theory that was not previously made FAIR. We used `Is derived from` to reference De Groot and Spiekerman's empirical cycle.
- Optionally, add *References* to related works in plain text. For example, we cite De Groot and Spiekerman in this field.

## Automating these Steps

R-users can use the `theorytools` package to partly automate the preceding steps, for example, using following code (see the package documentation for more information):

```
install.packages("theorytools")
library(theorytools)
# Use worcs to check if GitHub permissions are set:
library(worcs)
check_git()
check_github()
# Create the theory repository:
fair_theory(path = "c:/theoryfolder/empirical_cycle",
            title = "The Empirical Cycle",
            theory_file = "empirical_cycle.dot",
            remote_repo = "empirical_cycle",
            add_license = "cc0")
```
The first two lines install and load the `theorytools` package. Lines 4--6 use the `worcs` package to check whether git for version control is installed and a connection to github can be established to publish the theory. Line 8 calls the main entry function `fair_theory` for a local repository in a specified local `path` with a given `title` and the aforementioned dot `theory_file` containing the theory, uploading it to the user's remote github repository with name given in `remote_repo` and a given license specified in `add_license`.

@@

Several authors have reinterpreted De Groot's empirical cycle.
An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
For example, the DOT graph below implements Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle.
First, notice that the phases of the cycle have been renamed.
This change was not described in the paper.
If we assume that the labels are meant to illustrate the phases,
not substantially change the ontology,
then we can represent it by adding labels to the original DOT graph.
These labels suggest a focus on empirical psychology that was not present in De Groot's version.
Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
The DOT graph below shows our implementation of this version of the empirical cycle, by adding subgraphs.

```
digraph {

  subgraph cluster_discovery {
    label="Discovery";
    induction [label="New hypothesis"];
    deduction [label="New prediction"];
  }
  observation  [label="Old knowledge and old data"];
  subgraph cluster_justification {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

@@

}
```

The first author was inspired by De Groot as well,
but again specified the empirical cycle differently.
First, in De Groot's formulation, each stage describes a process.
This invites the question of what the concrete outcomes of these processes are.
In other words: what actually changes when going through the cycle, except the scholar's mind?
To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
<!-- In our representation, -->
<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
<!-- For example, the refutation of a hypothesized effect, -->
<!-- or the serendipitous observation of some pattern in data, -->
<!-- might be a reason to revise or construct theory. -->
<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
<!-- in the form of the inductive bias of methods used to perform the test, -->
<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->
Theory testing, as takes place in the "context of justification", can be said to involve mostly deductive reasoning.
Theory development and amendment, as takes place in the "context of discovery", involves primarily inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
However, deriving hypotheses from theory is not purely deductive
as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
Similarly, if we consider the claim that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
These alterations result in the following implementation of the empirical cycle:

```
digraph {

@@

}
```

The FAIR theory workflow documented above offers several concrete ways to make changes to a FAIR theory object.
If you start with an existing GitHub repository, and wish to make some changes to it,
this is commonly done by [creating a "branch"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-and-deleting-branches-within-your-repository).
A branch allows you to make non-destructive changes and can continue to exist in parallel to the main repository.
Thus, it is possible to have one main theory with several branches that each contain competing theories derived from it.
Figure \@ref(fig:figgithub-1) shows how our example `empirical_cycle` repository contains branches with Wagenmakers' and Van Lissa's implementations.
A branch can also be merged with the main branch, thus incorporating the changes it contains into the version of record.
If you wish to develop a new version of someone else's FAIR theory,
it is possible to ["fork their repository"](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo).
This creates a copy of their repository onto your GitHub account.
Both branches and forks can be compared and merged via "pull requests", which are essentially a request to incorporate the changes you have made.
Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.

```{r figgithub, out.width = "50%", fig.show='hold', fig.cap='FAIR Theories on GitHub', fig.subcap=c('Branches contain different versions of a theory', 'These versions can be compared'), fig.ncol = 2}
#par(mfrow=c(1,2))
knitr::include_graphics("git_branches.png")
knitr::include_graphics("git_compare_short.png")
```

@@

As uses of FAIR theory are best illustrated using tutorial examples,
the `theorytools` package contains several vignettes that showcase specific applications.
At the time of writing, the package includes a
vignette [introducing augmented Directed Acyclic Graphs (aDAGs)](https://cjvanlissa.github.io/theorytools/articles/augmented_dags.html) as a format for theory specification that meets the requirements of good psychological theory.
These aDAGs are X-interoperable for plotting (using `dagitty` and `tidySEM`),
for automatically selecting control variables, and for simulating data (using `theorytools`).
Another vignette describes how to take Self-Determination Theory [@langeSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].
Another vignette illustrates the use of FAIR theory for [covariate selection and causal inference](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].
More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
<!-- Some have argued that good theories involve *causal explanations* [@meehlAppraisingAmendingTheories1990]. -->
<!-- Explicit assumptions about the direction of causality allow one to perform causal inference even on cross-sectional data [-@pearlCausalDiagramsEmpirical1995]. -->
<!-- Any formal theory that is explicit about the direction of causality could thus be used to guide causal inference, -->
<!-- and could even be integrated into the analysis environment. -->

<!-- NOTES -->

<!-- * Theory is the vehicle of cumulative knowledge acquisition -->
<!-- * According to the empirical cycle, ideally, hypotheses are derived from theory, then tested in data, and theory is amended based on the resulting insights. When this cycle is regularly completed, theories become ever more veracious representations of social scientific phenomena. -->
<!-- * At present, there is concern over a theory crisis in the social sciences, which highlights that this system is not functioning as intended, and highlights the need for better theory. -->
<!-- * One source of potential improvements of theory methodology that has not been previously considered is computer science. -->
<!-- * The process of "iteratively improving" digital objects - in this case, computer code - is well understood. -->
<!-- * Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well. -->
<!-- * This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science. -->
<!-- * We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories -->
<!-- * FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory. -->
<!-- * FAIR theory allows for algorithmic derivation of hypotheses implied by the theory. -->
<!-- * FAIR theory enables integration testing: researchers can build a "test suite" of evidence that must be explainable by the theory, and any modifications of the theory must also pass the test suite. -->
<!-- * To illustrate FAIR theory's potential to accelerate cumulative knowledge acquisition, we present several tutorial examples, developed in collaboration with applied researchers across fields of social science. -->

>>>>

To concretize the FAIR principles,
we propose an applied workflow for making theory FAIR.
This conceptual workflow is software-agnostic,
but for R-users, most steps have been automated in the `theorytools` package.
We briefly introduce the conceptual workflow here, in terms of general principles that can be implemented using other open science infrastructures.
The R-package includes a living document with the [latest version of the workflow in tutorial format](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html),
using the specification and FAIRification of De Groot's "empirical cycle" as a guiding example.
Two other vignettes apply these processes to theories of greater substantive interest: Decy and Ryan's [*Self-Determination Theory*](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html) [@ryanSelfdeterminationTheoryFacilitation2000], and Morris' [*Tripartite Model*](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) of parental socialization of emotions in children [@morrisRoleFamilyContext2007].

Our approach builds upon existing successful open science infrastructures to the greatest possible extent.
At the time of writing (2024),
the integration of GitHub and Zenodo makes for a particularly user-friendly approach that meets *all* FAIR principles.
Zenodo and GitHub are both integrated with the Open Science Framework (OSF),
a popular platform in psychology.
Thus, it is possible to create a project page on the OSF to increase the visibility of a FAIR theory among users of that platform,
while the integration of the OSF with Zenodo and GitHub removes the need for maintaining the same information on multiple platforms.
Note that open science infrastructure is an area of active development, and as such,
workflows might change as new tools or databases are developed or existing tools and database change over time.

@@

If we compare this to the levels of theory formalization [@guestHowComputationalModeling2021],
it is defined at either the "theory" or "specification" level.
This theory can be FAIRified in its present form by archiving a text document containing five natural language statements.
However, it is also possible to increase the level of formalization,
and present an "implementation" of the theory as a directed graph (note: the implementation printed here is simplified; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details).
This fulfills criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
Given the cyclical nature of the conceptual model, such an implementation might look like this:

```
  induction;
  deduction;
  observation;
  test;
  evaluation;

  observation -> induction;
  induction -> deduction;
  deduction -> test;
  test -> evaluation;
  evaluation -> observation;
```

Note that the first part of the implementation constitutes an ontology - it specifies the entities comprised in the theory.
The second part of the implementation describes the flow of information from phase to phase.
Figure \@ref(fig:figecs)a) shows what this implementation looks like when plotted.
Regardless of which implementation we prefer - we can save it to a text file.
This text file is the "information artifact" containing our theory.

## 2. Create a Theory Archive

To help meet the Interoperability and Reusability criteria,
it is important to bundle the theory file with additional files containing documentation.
Firstly, add a README.md file with instructions for future users of your theory.
The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) with guidance on writing a README file for theory.
Secondly, add a LICENSE file with the legal conditions for reuse.
As explained previously, we recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
Place all files in a folder together - this folder will become the theory archive.

## 3. Version Control the Archive

To track all changes to our theory, the previously created folder can be version controlled.
[Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) is well-suited for this purpose.
To invite reuse and collaboration, it is further recommended to create a backup in the cloud on a service like [GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/).

@@

Aside from tracking changes, it is also important to retain a permanent archive of major versions of a theory,
so that they can be referenced by future research in perpetuity.
Zenodo is a FAIR compliant repository for archiving information artifacts, with GitHub integration.

@@

Once archived, documenting a FAIR theory with relevant metadata improves its Findability.
We recommend using a standardized metadata schema, such as DataCite [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
Within this schema, set the *resource type* to `Model`, add the words `FAIR theory:` to the title so that sentient readers to recognize the work as a FAIR theory, and add `fairtheory` as a keyword to aid search engine indexation. Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
As mentioned before, [these five steps are partly automated for R-users](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html).
See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle.

@@

An important advantage of FAIR theory is that we can implement different versions of a theory, compare them, and document their cross-relationships.
We can take work that has been done before - in this case, the repository created above, and create an independent copy that we can modify as we wish, while retaining cross-references to the original.
Elaborating on our running example, several authors have reinterpreted De Groot's empirical cycle.
For example, Wagenmakers and colleagues' [-@wagenmakersCreativityVerificationCyclePsychological2018] interpretation of the empirical cycle differs from De Groot's in several ways.
First, the phases of the cycle were renamed, and this change was not described in the paper.
Assuming that Wagenmakers' new labels are meant to illustrate the phases,
not substantially change the ontology,
we could incorporate this change by adding labels to the original ontology.
These labels suggest a focus on empirical psychology that was not present in De Groot's version.
Furthermore, the label "knowledge accumulation" invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
As this lack of cumulative knowledge acquisition appears to be precisely where contemporary research practice falls short, this ambiguity invites further improvement of the theory.
The authors explicitly mention a second change: *"We added the Whewell-Peirce-Reichenbach distinction between the context of discovery and the context of justification"* (p. 423).
We could implement this change to the original implementation by grouping the respective phases of the cycle;
a minor and tractable change.

```
  {
    label="Discovery";
    induction [label="New hypothesis"];
    deduction [label="New prediction"];
  }
  observation  [label="Old knowledge and old data"];
  {
    label="Justification";
    test [label="Test on new data"];
    evaluation;
  }

@@

```

The first author was inspired by De Groot as well,
but again specified the empirical cycle differently.
First, in De Groot's formulation, each stage describes a process.
This invites the question of what the concrete outcomes of these processes are.
In other words: what actually changes when going through the cycle, except the scholar's mind?
To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs)c).
Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
Theory testing, as takes place in the "context of justification", can be said to involve mostly deductive reasoning.
Theory development and amendment, as takes place in the "context of discovery", involves primarily inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
However, deriving hypotheses from theory is not purely deductive
as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances.
Similarly, if we consider the claim that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
These alterations result in the following implementation of the empirical cycle:

```

@@

```

The FAIR theory workflow offers concrete ways to make changes to a FAIR theory object and to compare different versions,
as explained in [this vignette](ADD ME).
For example, Figure \@ref(fig:figgithub-2) shows a comparison of the original empirical cycle by De Groot, and the lead author's implementation.

```{r figgithub, out.width = "50%", fig.show='hold', fig.cap='FAIR Theories on GitHub', fig.subcap=c('Different versions of a theory are archived as distinct branches', 'These versions can be explicitly compared'), fig.ncol = 2}
#par(mfrow=c(1,2))
knitr::include_graphics("git_branches.png")
knitr::include_graphics("git_compare_short.png")
```

@@

As uses of FAIR theory are best illustrated using tutorial examples,
the `theorytools` package contains several vignettes that showcase specific applications.
At the time of writing, the package includes a
vignette [introducing augmented Directed Acyclic Graphs (aDAGs)](https://cjvanlissa.github.io/theorytools/articles/augmented_dags.html) as a format for theory specification that meets the requirements of good psychological theory.
These aDAGs are X-interoperable for plotting (using `dagitty` and `tidySEM`),
for automatically selecting control variables, and for simulating data (using `theorytools`).
Another vignette describes how to take Self-Determination Theory [@langeSelfDeterminationTheory2012], a theory originally represented as prose, and [specify it as a FAIR aDAG](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html).
Another vignette describes how to take the Dunning-Krüger effect and [specify it as a FAIR mathematical formula](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html) [@feldEstimatingRelationshipSkill2017].
Another vignette illustrates the use of FAIR theory for [covariate selection and causal inference](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html) [@pearlCausalDiagramsEmpirical1995].
More vignettes may be added over time, and users are encouraged to submit their own reproducible examples as package vignettes.
